% $Date: 2026/05/02 16:49:56 $
% This template file is public domain.
%
% TUGboat class documentation is at:
%   texdoc tugboat
% or
%   https://texdoc.org/serve/tugboat/0
% or
%   https://mirrors.ctan.org/macros/latex/contrib/tugboat/ltubguid.pdf

\documentclass{ltugboat}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{microtype}
\usepackage[hidelinks]{hyperref}

\title{\LaTeX\ for Elegant Humanities Handouts in Conferences and Classrooms}

% repeat info for each author; comment out items that don't apply.
\author{Corey J. Stephan}
\address{2126 Berkhsire Elm Street \\ Katy, Texas, USA 77493 \\}
\netaddress{corey.stephan@stthom.edu}
\personalURL{https://coreystephan.com}
%\ORCID{0}

% Please state if you'd like to receive a physical copy of the TUGboat
% issue or if electronic access suffices.  If you want a physical issue,
% please include the mailing address we should use, as a comment if
% you prefer it not be printed.

% Physical copy, please, mailed here:
%
% Corey Stephan
% 2126 Berkshire Elm Street
% Katy, Texas, USA 77493
%

% Corey Stephan's package declarations
\usepackage{polyglossia}
\usepackage{parallel}
\usepackage{hologo}
\usepackage[olddefault]{fontsetup}


\begin{document}
	
% Corey Stephan's linguistic declarations
\setdefaultlanguage{english}
\setotherlanguage[variant=ancient]{greek}
	
\maketitle

\begin{abstract}
{\LaTeX\ has been popular throughout the mathematical and physical sciences for decades. \LaTeX's utility for scholars in humanities disciplines, however, tends to be overlooked. In this tutorial, I introduce the use of \LaTeX\ for a common task in humanities scholarship, namely, typesetting texts with unusual line length and other organizational needs, such as poetry and translations. The specific example is preparing a handout for a historical conference or class presentation with excerpts of ancient Greek and parallel English translations. Ultimately, I aim to inspire colleagues in religious, historical, philosophical, and related areas to think of \LaTeX\ as having a wide range of professional applications even inside these scholarly fields in which \LaTeX\ is not widespread.}
\end{abstract}

\section{Introduction}

The world's best established software system for typesetting and document formatting, \LaTeX\ has been popular throughout mathematical and physical sciences for decades. The utility of \LaTeX\ for humanities disciplines, however, tends to be overlooked. \LaTeX\ can be used for typesetting in most instances in which a scholar of humanities needs to present materials. Whether one works in history, philosophy, literature, or some other discipline in which \LaTeX\ is not usually employed, \LaTeX\ is nevertheless worth the investment of time to learn.

Granted, in disciplines outside the mathematical and physical sciences, such as my own area of historical theology (technically, divinities rather than humanities), editors expect us to prepare our materials for publications with bug-for-bug compatibility with Microsoft Word for back-and-forth markup. In practice, this expectation requires us to use word processors, such as LibreOffice Writer. A word processor presents material in a What You See Is What You Get (WYSIWYG, ``whizzy-wig'') style. While one writes, one is simultaneously thinking about and working on formatting.

In a typesetting language, such as \LaTeX\ (or \ConTeXt\ or Typst, either of which a scholar might also choose for various reasons), one is only concerned about the structure of what one is writing. One declares that specific characters will be in certain categories of writing `things' (heading, paragraph, footnote, accent type, typeface, and so on), and one allows the program to compile one's writing as `code' into a tidy final document \Dash every time. Preparing a \textit{curriculum vitae}, conference handout, or other document with unusual typesetting requirements becomes about the writing itself rather than the formatting.

To learn \LaTeX\ as a scholar outside the disciplines in which \LaTeX\ is the norm, I recommend working through one of two community-built, libre guidebooks. The first, which I used to learn \LaTeX, is \titleref{The Not So Short Introduction to \LaTeX} (\url{github.com/oetiker/lshort}). The second, which a colleague of mine recommends, is \titleref{Modern \LaTeX} (\url{github.com/mrkline/modern-latex}). Either guidebook may be used to begin to learn how to typeset cover letters, ancient languages, and other materials that are relevant across the humanities \Dash without the front-loading of mathematics and physical sciences that one finds in other general introductions.

To show how thinking about digital typesetting in terms of document structure rather than final format can be of practical use in the humanities, here I present a tutorial in which I introduce the use of \LaTeX\ for a common task in humanities scholarship, namely, typesetting texts with unusual line length and other organizational needs. The thought experiment is that of determining how to prepare a handout for a historical conference or class presentation with perfectly organized excerpts of a Byzantine Greek text and an original English translation in parallel.

\section{Simplest parallel text in \LaTeX}

There are many ways to set parallel text in \LaTeX. The simplest is to use the community package called \texttt{parallel}, which is included in \hologo{TeXLive}, \hologo{MiKTeX}, and other \LaTeX\ distrubtions. For full documentation, visit \url{https://ctan.org/pkg/parallel}.

To start using \texttt{parallel} in any \LaTeX\ document, include the following in the preamble:
{\small \begin{verbatim}
\usepackage{parallel}
\end{verbatim}}
For example, here is an abbreviation of the preamble to the Beamer file for my talk at \TUG\ 2026 (\url{github.com/historical-theology/latex-talks/}) with only the most important entries, including \texttt{parallel}:
{\small \begin{verbatim}
\documentclass[aspectratio=169]{beamer}
\usetheme{Goettingen}

\usepackage{parallel}

\title{\LaTeX\ for Elegant Humanities Handouts 
			\\ in Conferences and Classrooms }
\author{Corey Stephan, Ph.D.}

\begin{document}
\end{verbatim}}

This is the basic format for parallel text set with the package \texttt{parallel}:

{\small \begin{verbatim}
\begin{Parallel}{}<left space>}{<right space>}
\ParallelLText{<left text row 1>}
\ParallelRText{<right text row 1>}
\ParallelPar
\ParallelLText{<left text row 2>}
[and so on]
\end{Parallel}	
\end{verbatim}}
I (a theologian) might typeset the opening words of the Christian prayer \textit{Ave Maria} in their traditional Latin with parallel English like so:

{\small \begin{verbatim}
\begin{Parallel}{0em}{0em}
	\ParallelLText{Ave Maria,}
	\ParallelRText{Hail Mary,}
		\ParallelPar
	\ParallelLText{gratia plena \ldots}
	\ParallelRText{full of grace \ldots}
		\ParallelPar
\end{Parallel}	
\end{verbatim}}


Here is the result:

\medskip
\begin{minipage}{.4\textwidth}
\begin{Parallel}{0em}{0em}
\ParallelLText{Ave Maria,}
\ParallelRText{Hail Mary,}
\ParallelPar
\ParallelLText{gratia plena \ldots}
\ParallelRText{full of grace \ldots}
\ParallelPar
\end{Parallel}
\end{minipage}

\section{Advanced parallel text in \LaTeX}

For advanced scholarly setting of parallel text in \LaTeX, especially long-form (e.g. a critical edition with accompanying translation), one ought to learn the two interconnected and frequently updated community packages \texttt{reledpar} (for parallel typesetting specifically) and \texttt{reledmac} (for typesetting academic editions in general), which are bundled with \hologo{TeXLive} and other \LaTeX\ distributions.

\texttt{reledpar} and \texttt{reledmac} are purpose-built for long-form academic parallel texts. These two packages are meant to replace older methods of academic parallel text preparation in \LaTeX\ that one might find in Web searches. For full documentation, visit \url{ctan.org/pkg/reledpar} and \url{ctan.org/pkg/reledmac}.

\section{Setting non-Latin characters in \LaTeX}

The most important packages for handling text outside Latin alphabets are \texttt{babel} and \texttt{polyglossia}, which are included in \hologo{TeXLive} and other \LaTeX\ distributions. Although it is possible to convert a project from one of these metapackages to the other, a scholar ought to invest the time to choose carefully between either of these two at the start of a long-form project.

\subsection{babel vs. polyglossia}

In \titleref{The \LaTeX\ Companion: 3rd Edition} (Addison-Wesley, 2023), Frank Mittelbach and Ulrike Fischer suggest that despite the fact that \texttt{babel} is older than \texttt{polyglossia}, \texttt{babel} generally is preferrable: ``Unless you are in a special situation where a language or script is notably better supported by \texttt{polyglossia}, the recommendation is to use \texttt{babel} with all engines'' (entry 13.7.2, ``Polyglossia''). Since \texttt{babel} technically supports the same languages as \texttt{polyglossia}, a situation in which one would need \texttt{polyglossia} over \texttt{babel} will be rare. Yet, a scholar working with ancient languages might have such a rare situation.

In (old) Web forum discussions, one might read that \texttt{babel} will work better with \LuaLaTeX, whereas \texttt{polyglossia} will work better with \XeLaTeX. In 2026, however, one may choose either of the two mainline \LaTeX\ engines that are built with full Unicode support, that is, \XeLaTeX\ or \LuaLaTeX, for \texttt{babel} or \texttt{polyglossia}. Results might vary, however, depending on the specifics of a given project.

To make an informed choice against a specific set of requirements for a given academic undertaking, one should explore each package's documentation at \url{ctan.org/pkg/babel} and \url{ctan.org/pkg/polyglossia}.

To learn \texttt{babel}, I recommend Chapter 13: ``Localizing documents'' in \titleref{The \LaTeX\ Companion: 3rd Edition} by Mittelbach and Fischer, which succinctly introduces localization in \LaTeX\ via \texttt{babel} in a practical, didactic (teaching) style.

To learn \texttt{polyglossia}, I recommend the package's official documentation in \CTAN.

\section{Starting a project with \texttt{polyglossia}}

Here, I will model \texttt{polyglossia}, which is what I happen to be familiar with using for polytonic (ancient and Byzantine) Greek.

First, include the package in the preamble:
{\small \begin{verbatim}
\usepackage{polyglossia}
\end{verbatim} 
}
Then, after the beginning of the document but before actual type, declare the languages and their variants. For example, here is this .tex file's \texttt{polyglossia} declarations:

{\small \begin{verbatim}
\begin{document}
	\setdefaultlanguage{english}
	\setotherlanguage[variant=ancient]{greek}
\end{verbatim}}

One is able to use one of the `other' declared languages (in this case, ancient Greek) by toggling it when needed.

A declared foreign or ancient language may be toggled for a long excerpt like so:

{\small \begin{verbatim}
\begin{greek}
	<greek text>
	...
\end{greek}
\end{verbatim}}
%
For example, this is how one could set the famous opening line of John's Gospel:
{\small \begin{verbatim}
\begin{greek}
ἐν ἀρχῇ ἦν ὁ λόγος, \\
καὶ ὁ λόγος ἦν πρὸς τὸν θεόν, \\
καὶ θεὸς ἦν ὁ λόγος. \\
\end{greek}
\end{verbatim}} 
%
Here is the clean result: \\
\begin{greek}ἐν ἀρχῇ ἦν ὁ λόγος, \\
καὶ ὁ λόγος ἦν πρὸς τὸν θεόν, \\
καὶ θεὸς ἦν ὁ λόγος. \\
\end{greek} 
%
For another example, here are the first words of Homer's \titleref{The Iliad}:
%
{\small \begin{verbatim}
\begin{greek}
μῆνιν ἄειδε θεὰ Πηληϊάδεω Ἀχιλῆος \\
οὐλομένην, ἣ μυρί' Ἀχαιοῖς ἄλγε' ἔθηκε,
\end{greek}
\end{verbatim}} 
%
Here is the equally clean result: \\
\begin{greek}μῆνιν ἄειδε θεὰ Πηληϊάδεω Ἀχιλῆος \\
οὐλομένην, ἣ μυρί' Ἀχαιοῖς ἄλγε' ἔθηκε,
\end{greek} 


A declared language may be toggled for a single word or line like so:

{\small \begin{verbatim}
\text<language>{<text in language>}
\end{verbatim}}
%
This might be used to insert a keyword, e.g. \break 
{\textgreek{ὁ λόγος}} (produced with \verb|\textgreek{ὁ λόγος}|), inside running prose.

\section{Combining \texttt{parallel} and \texttt{polyglossia}}



\makesignature
\end{document}
