% Main Document Settings
\documentclass{beamer}
\usetheme{Goettingen}

% Enable polyglossia and alter default fonts to display polytonic Greek
\usepackage{polyglossia}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSerif}
\setmonofont{FreeMono}

% Allow fontawesome icons, if desired
\usepackage{fontawesome}

% Allow code quotations
\usepackage{listings}

% Demonstrational packages for parallel typesetting
\usepackage{parallel}
%\usepackage{reledmac}
%\usepackage{reledpar}

\title{\LaTeX\ for Medieval Primary Source Handouts}
\author{Corey Stephan, Ph.D. \break \href{https://www.coreystephan.com}{coreystephan.com}}

\begin{document}
\setdefaultlanguage{english}
\setotherlanguage[variant=ancient]{greek}
	
\begin{frame}[plain]
    \maketitle
\end{frame}

\begin{frame}
    \frametitle[Everything]{\LaTeX\ for Everything}
The world's best established software system for typesetting and document formatting, \LaTeX\ has been popular throughout the mathematical and physical sciences for decades. The utility of \LaTeX\ for scholars in humanities disciplines, however, tends to be overlooked. \break

\LaTeX\ can be used for typesetting in most instances in which a scholar of humanities needs to present our materials. \break

I prepared these slides in \LaTeX\ -- with less frustration than would have been possible in graphical presentation program (e.g. LibreOffice Present) -- as 1 file: \break

\small{\href{https://github.com/historical-theology/latex-talks/blob/main/icms/medievalist-latex.tex}{github.com/historical-theology/latex-talks/blog/main/icms/medievalist-latex.tex}}
\end{frame}

\begin{frame}
	\frametitle[Tutorial]{This Tutorial}
In this tutorial, I will introduce attendees to \LaTeX, as well as to the use of \LaTeX\ for a common task in medievalist scholarship, namely, typesetting texts with unusual line length and other organizational needs, such as primary texts with our original translations. \break

The specific `thought experiment' will be that of determining how to prepare a handout for a historical conference or class presentation with perfectly organized excerpts of a Byzantine Greek text and an original English translation in parallel. \break
\end{frame}

\begin{frame}
	\frametitle[\LaTeX]{What is \LaTeX ?}
It might be best to start to understand what \LaTeX\ \emph{is} by understanding what it is \emph{not}. \break

A word processor, such as LibreOffice Writer or Microsoft Word, presents material in a (W)hat (Y)ou (S)ee (I)s (W)hat (Y)ou (G)et (WYSIWYG, ``whizzy-wig'') style. While one writes, one is simultaneously thinking about and working on formatting. \break

In a typesetting language, such as \LaTeX\ (or ConTeXt or Typst, among others), one is only concerned about the \emph{structure} of what one is writing. One declares that specific characters will be in certain categories of writing `things' (heading, paragraph, footnote, accent type, typeface, etc.), and one allows the program to compile one's writing as `code' into a tidy final document -- every time.
\end{frame}

\begin{frame}
	\frametitle[Starting \LaTeX]{Getting Started with \LaTeX\ in GNU/Linux}
\begin{itemize}
	\item \href{https://www.latex-project.org/}{latex-project.org}
		\begin{itemize}
			\item official recommendations for \LaTeX\ \textbf{distributions}
			\item \LaTeX\ \textbf{distributions} are community-maintained bundles of closely interrelated software, including the \TeX\ backend, \LaTeX\ itself, and hundreds to thousands of packages that extend \LaTeX's functionality
		\end{itemize}
	\item *BSD or GNU/Linux? 
	\textbf{TeXLive}
	\item MacOS? 
	\textbf{MacTeX}
	\item Microsoft Windows? 
\textbf{TeXLive} or \textbf{MiKTeX}
\end{itemize}
\end{frame}

\begin{frame}
	TexLive is available in the official package repositories of most *BSD and GNU/Linux operating systems. To install the complete TeXLive distribution (which the \LaTeX\ project recommends for newcomers) on one of the most popular libre Unix-live operating systems, one could type: \break
	
	Debian/Ubuntu \break
	\texttt{apt install texlive-full} \break
	
	Fedora \break
	\texttt{dnf install texlive-scheme-full} \break
	
	Arch \break
	\texttt{yay -S texlive-meta} \break
	
	FreeBSD \break
	\texttt{pkg install texlive-full}
\end{frame}

\begin{frame}
	
	2 community-built, libre guidebooks \break
	
	I recommend \break \emph{The Not So Short Introduction to \LaTeX} \break
	\href{https://github.com/oetiker/lshort}{github.com/oetiker/lshort} \break
	
	Dr. Hayes helpfully recommends \break \emph{Modern \LaTeX} \break
	\href{https://github.com/mrkline/modern-latex}{github.com/mrkline/modern-latex}
\end{frame}

\begin{frame}[fragile]
	\frametitle[Simple Parallel Text]{Simple Setting of Parallel Text in \LaTeX}
There are many ways to set parallel text in \LaTeX. \break

The \emph{simplest} is to use the community package called \texttt{parallel}, which is included in TeXLive, MiKTeX, and other \LaTeX\ distrubtions. For full documentation, visit: \break 

\href{https://ctan.org/pkg/parallel}{ctan.org/pkg/parallel} \break

To start using \texttt{parallel} in any \LaTeX\ document, simply include the following in the preamble: \break
 
\verb|\usepackage{parallel}| \break
\end{frame}

\begin{frame}[fragile]
Here is an abbreviated form of the preamble to this Beamer (\LaTeX\ presentation class) file with the most important entries, including \texttt{parallel}: \break

{\scriptsize \begin{lstlisting}
\documentclass{beamer}
\usetheme{Goettingen}

\usepackage{listings}
\usepackage{parallel}

\title{\LaTeX\ for Medieval Primary Source Handouts}
\author{Corey Stephan, Ph.D. \break 
\href{https://www.coreystephan.com}{coreystephan.com}}

\begin{document}
\end{lstlisting}}
\end{frame}

\begin{frame}[fragile]
	\frametitle[Using parallel]{Using the \texttt{parallel} package}

Basic Format

{\scriptsize \begin{lstlisting}
\begin{Parallel}{}<left adjustment>}{<right adjustment>}
		\ParallelLText{<left text row 1>}
		\ParallelRText{<right text row 1>}
			\ParallelPar
		\ParallelLText{<left text row 2>}
	...	
\end{Parallel}	
\end{lstlisting}}

Example
	
{\scriptsize \begin{lstlisting}
\begin{Parallel}{0mm}{0mm}
	\ParallelLText{Ave Maria,}
	\ParallelRText{Hail Mary,}
		\ParallelPar
	\ParallelLText{gratia plena,}
	\ParallelRText{full of grace,}
\end{Parallel}	
		\end{lstlisting}}

\begin{Parallel}{0mm}{0mm}
\ParallelLText{Ave Maria,}
\ParallelRText{Hail Mary,}
\ParallelPar
\ParallelLText{gratia plena,}
\ParallelRText{full of grace,}
\end{Parallel}	

\end{frame}

\begin{frame}[fragile]
	\frametitle[Advanced Parallel Text]{Advanced Setting of Parallel Text in \LaTeX}
	
For \emph{advanced} scholarly setting of parallel text in \LaTeX, one ought to learn the 2 interconnected and frequently updated community packages \texttt{reledpar} (for parallel typesetting specifically) and \texttt{reledmac} (for typesetting academic editions in general), which are bundled with TeXLive and other \LaTeX\ distributions. \break

These 2 packages are specifically designed for long-form academic parallel texts, and they are meant to replace older methods of academic parallel text preparation in \LaTeX\ that you might find in Web searches. \break

For full documentation, visit: \break
\href{https://ctan.org/pkg/reledpar}{ctan.org/pkg/reledpar} \break
\href{https://ctan.org/pkg/reledmac}{ctan.org/pkg/reledmac} \break

Remember them in the preamble to your document:

{\scriptsize \begin{lstlisting}
\usepackage{reledpar}
\usepackage{reledmac}
\end{lstlisting}}

\end{frame}

\begin{frame}[fragile]
	\frametitle[Non-Latin Text]{Setting of Non-Latin Text in \LaTeX}

The most important package for handling text outside of Latin alphabets is the community package \texttt{polyglossia}, which is (as you probably have guessed) included in TeXLive and other \LaTeX\ distributions. \break

This fresh and continuously updated package is meant to replace older methods of non-Latin text entry that you might find in Web searches, notably \texttt{babel}, which ought to be set aside in 2026 (except in legacy projects that already rely on it). \break

For full documentation, visit: \break
\href{https://www.ctan.org/pkg/polyglossia}{ctan.org/pkg/polyglossia} \break

\end{frame}

\begin{frame}[fragile]
	\frametitle[Using polyglossia]{Starting a Project with \texttt{polyglossia}}

Remember the package in the preamble:

\begin{lstlisting}
\usepackage{polyglossia}
\end{lstlisting} 

\vspace{1em}

Then, \emph{after} the beginning of the document but \emph{before} actual type, declare the languages and their variants. For example, here is this .tex file's polyglossia declarations:

\begin{lstlisting}
\begin{document}
\setdefaultlanguage{english}
\setotherlanguage[variant=ancient]{greek}
\end{lstlisting}

\end{frame}

\begin{frame}[fragile]	
	
Then, when using one of the `other' declared languages, toggle it when needed, either for long excerpts as
	
{\scriptsize \begin{lstlisting}
	\begin{greek}
		<greek text>
		...
	\end{greek}
\end{lstlisting}}
	
or for single lines as
	
{\scriptsize \begin{lstlisting}
	\text<language>{<text in language>}
	e.g.
	\textgreek{<Greek text>}
\end{lstlisting}} \break

The former might be used to set apart long quotations, e.g. \break

{\scriptsize \begin{greek}
ἐν ἀρχῇ ἦν ὁ λόγος, \\
καὶ ὁ λόγος ἦν πρὸς τὸν θεόν, \\
καὶ θεὸς ἦν ὁ λόγος. \\
\end{greek}}

\break \vspace{1em}

while latter might be used to insert a keyword in another language, e.g. {\textgreek{ὁ λόγος}}, in the midst of prose.
	
\end{frame}

\begin{frame}[fragile]
	\frametitle[Combining parallel and polyglossia]{Combining \texttt{parallel} and \texttt{polyglossia}}

{\small The simplest way of showing multilingual text in parallel is} \break

{\scriptsize \begin{lstlisting}
\begin{Parallel}{0mm}{0mm}
	\text<language>{\ParallelLText{<ancient text 1>}}
	\ParallelRText{<modern text>}
		\ParallelPar
	\text<language>{\ParallelLText{<ancient text 2>}}
	\ParallelRText{<modern text 2>}
\end{Parallel}	
\end{lstlisting}} \break

\vspace{1em}

{\small but since this is tedious for more than a few lines, I remind everyone of \texttt{reledpar} and \texttt{reledmac} for longer and/or more complicated parallel text displays.} \break

\end{frame}

\begin{frame}[fragile]

{\small Canon in Honor of Thomas Aquinas, Ode 3}\footnote{Raffaele Cantarella, ed., “Canone greco inedito di Giuseppe vescovo di Methone in onore di San Tommaso d’Aquino,” \emph{Archivum Fratrum Praedicatorum} 12 (1934): 145--85 at 155--58.} \\
{\small Translated by Corey Stephan, Ph.D.} \break

{\scriptsize \begin{Parallel}{0mm}{0mm}
	\textgreek{\ParallelLText{Νῦν συγχορεύεις ἀληθῶς}}
	\ParallelRText{Now, truly, you celebrate together}
		\ParallelPar
	\textgreek{\ParallelLText{σὺν τοῖς λοιποῖς θεολόγοις}}
	\ParallelRText{with the rest of the theologians}
		\ParallelPar
	\textgreek{\ParallelLText{καὶ τοῖς ἄλλοις θεολήπτοις ἀνδράσι,}}
	\ParallelRText{and the other divinely inspired men}
		\ParallelPar
	\ParallelRText{with the rest of the theologians}
	\textgreek{\ParallelLText{περὶ τὸν θρόνον Χριστοῦ}}
	\ParallelRText{around the throne of Christ,}
		\ParallelPar
	\textgreek{\ParallelLText{τὸν ἱερὸν ἐξᾴδοντες}}
	\ParallelRText{singing the sacred}
		\ParallelPar
	\textgreek{\ParallelLText{καὶ τρισάγιον ὕμνον}}
	\ParallelRText{and Thrice-Holy Hymn}
		\ParallelPar
	\textgreek{\ParallelLText{σὺν ταῖς ἁγίαις δυνάμεσιν.}}
	\ParallelRText{with the holy powers.}
		\ParallelPar
	\vspace{1em}
	\textgreek{\ParallelLText{Ἔνθα πατέρων ὁ χορὸς}}
	\ParallelRText{Where there is the chorus of Fathers}
		\ParallelPar
	\textgreek{\ParallelLText{καὶ ἱερῶν θεολόγων}}
	\ParallelRText{and holy theologians,}
		\ParallelPar
	\textgreek{\ParallelLText{ὡς ὑψίνους καὶ λεπτὸς θεόλογος}}
	\ParallelRText{as a high-minded and subtle theologian}
		\ParallelPar
	\textgreek{\ParallelLText{κατεσκήνωσας, Θωμᾶ,}}
	\ParallelRText{you do abide, Thomas,}
		\ParallelPar
	\textgreek{\ParallelLText{σὺν τούτοις αὐλιζόμενος,}}
	\ParallelRText{dwelling with them,}
		\ParallelPar
	\textgreek{\ParallelLText{θεολογῶν ἀπαύστως}}
	\ParallelRText{unceasingly theologizing,}
		\ParallelPar
	\textgreek{\ParallelLText{τριάδα, πάτερ, τὴν ἄκτιστον.}}
	\ParallelRText{O Father, about the Uncreated Trinity.}
		\ParallelPar
\end{Parallel}}	

\end{frame}

\begin{frame}
	\frametitle[Other \LaTeX\ software]{Other \LaTeX\ Tools for Scholars}

\textbf{BibLaTeX} -- comprehensive citation and bibliography management, capable of being interlinked with a variety of citation managers (Zotero, JabRef, etc.) \break

\textbf{TeXStudio} and other graphical front-ends -- LaTeX can be written in any plain text editor, but a designated LaTeX writing environment improves efficiency and helps correct code errors \break

Designated typeface packages, e.g. \textbf{Serto} for Syriac \break

\textbf{microtype} for small adjustments to type, such as improved automatic hyphenation (a personal favorite)

\end{frame}

\end{document}
